---
layout: page
title: R for Data Analysis
subtitle: Data Analysis
minutes: 30
---

```{r, include=FALSE}
source("tools/chunk-options.R")
opts_chunk$set(fig.path = "fig/09-vectorisation-")
# Silently load in the data so the rest of the lesson works
hh <- read.table(file="data/city__table__households.csv", header=TRUE, sep=",")
cities <- read.table(file="data/cities.csv", header=TRUE, sep=",", stringsAsFactors=FALSE)
counties <- read.table(file="data/counties.csv", header=TRUE, sep=",", stringsAsFactors=FALSE)
hh <- merge(hh, cities, by="city_id", all=TRUE)
hh <- merge(hh, counties, by="county_id", all=TRUE)
colnames(hh)[3:6] <- c("hh10", "hh20", "hh30", "hh40")
pierce <- subset(hh, county_name == "Pierce")
```

> ## Learning Objectives {.objectives}
>
> * To be able to summarize and visualize data.
>

## Data manipulation

Let's try vectorisation on the `hh40` column of the `hh` dataset.
First remove rows with NA. Then make a new column in the `hh` data frame that
contains households in units of thousands of people.
```{r}
hh <- na.omit(hh)
hh$hh40inT <- hh$hh40 / 1000
head(hh)
```
Create a log transformation:
```{r}
hhlog <- hh
hhlog[,2:6] <- log(hhlog[,2:6])
head(hhlog)
```

Create a dataset of differences:
```{r}
hhdif <- cbind(data.frame(city_id=hh$city_id), hh[,3:5] - hh[,2:4])
head(hhdif)
summary(hhdif)
```

## Scatter plot, histogram, boxplot

Plot results:
```{r}
plot(hh$hh10, hh$hh40, xlab="2010", ylab="2040")
abline(0,1)
plot(hh$hh10, hh$hh40, xlab="2010", ylab="2040", log="xy")
abline(0,1)
hist(hh$hh40 - hh$hh10, main="Change 2040-2010")
abline(v=0, lwd=2, col="red")
```
Investigate records with negative change between 2040 and 2010:
```{r}
neg <- hh[hh$hh40 - hh$hh10 < 0, ]
dim(neg)
plot(neg$hh10, neg$hh40, xlab="2010", ylab="2040", log="xy")
text(neg$hh10, neg$hh40, labels=neg$city_name, pos=3)
abline(0,1)
```

Let's bring in a dataset with multiple land use indicators from 2010, merge it with city names and omit records with missing values:
```{r}
lu <- read.table(file="data/city__dataset_table__lu_indicators__2010.tab", header=TRUE, sep="\t")
lu <- merge(lu, cities, by="city_id", all=TRUE)
lu <- na.omit(lu)
summary(lu)
```
Say we want to look at the distribution of the average household sizes by counties. First, create a column for household size:  
```{r}
lu <- cbind(lu, hh_size=lu$population/lu$households)
head(lu[order(lu$hh_size),])
tail(lu[order(lu$hh_size),])
```
Second, create a histogram for all cities, and a box plot by counties:
```{r}
hist(lu$hh_size)
boxplot(hh_size ~ county_id, lu)
```

## Simple regression
Here is a simple regression for our lu dataset:
```{r}
plot(jobs ~ non_res_sqft, lu)
fit <- lm(jobs ~ non_res_sqft, lu)
summary(fit)
abline(fit)
cor(lu$jobs, lu$non_res_sqft)
```
Try to add another variable:
```{r}
summary(lm(jobs ~ non_res_sqft + population, lu))
```

Or remove the intercept:
```{r}
fit <- lm(jobs ~ -1 + non_res_sqft, lu)
summary(fit)
names(fit)
```

## Saving plots
You can save a plot from within RStudio using the 'Export' button
in the 'Plot' window. This will give you the option of saving as a
.pdf or as .png, .jpg or other image formats.

Sometimes you will want to save plots without creating them in the
'Plot' window first. Perhaps you want to make a pdf document with
multiple pages: each one a different plot, for example. Or perhaps
you're looping through multiple subsets of a file, plotting data from
each subset, and you want to save each plot, but obviously can't stop
the loop to click 'Export' for each one.

In this case you can use a more flexible approach. The function
`pdf` creates a new pdf device. You can control the size and resolution
using the arguments to this function.

```{r, eval=FALSE}
pdf("myplot.pdf", width=12, height=4)
par(mfrow=c(1,2))
hist(lu$hh_size)
boxplot(hh_size ~ county_id, lu)
# You then have to make sure to turn off the pdf device!
dev.off()
```
Open up this document and have a look.

The commands `jpeg`, `png` etc. are used similarly to produce
documents in different formats.

