---
layout: page
title: R for Data Analysis
subtitle: Creating functions
minutes: 45
---

```{r, include=FALSE}
source("tools/chunk-options.R")
# Silently load in the data so the rest of the lesson works
lu <- read.table(file="data/city__dataset_table__lu_indicators__2010.tab", header=TRUE, sep="\t")
cities <- read.table(file="data/cities.csv", header=TRUE, sep=",", stringsAsFactors=FALSE)
lu <- merge(lu, cities, by="city_id", all=TRUE)
lu <- na.omit(lu)
hh <- read.table(file="data/city__table__households.csv", header=TRUE, sep=",")
cities <- read.table(file="data/cities.csv", header=TRUE, sep=",", stringsAsFactors=FALSE)
counties <- read.table(file="data/counties.csv", header=TRUE, sep=",", stringsAsFactors=FALSE)
hh <- merge(hh, cities, by="city_id", all=TRUE)
hh <- merge(hh, counties, by="county_id", all=TRUE)
colnames(hh)[3:6] <- c("hh10", "hh20", "hh30", "hh40")
```

> ## Learning Objectives {.objectives}
>
> * Define a function that takes arguments.
> * Return a value from a function.
> * Test a function.
> * Set default values for function arguments.
> * Explain why we should divide programs into small, single-purpose functions.
>


In this lesson, we'll learn how to write a function so that we can repeat
several operations with a single command.

> ## What is a function? {.callout}
>
> Functions gather a sequence of operations into a whole, preserving it for ongoing use. Functions provide:
>
> * a name we can remember and invoke it by
> * relief from the need to remember the individual operations
> * a defined set of inputs and expected outputs
> * rich connections to the larger programming environment
>


## Defining a function

Let's open a new R script file  and call it functions-lesson.R.

```{r}
sqft.to.acres <- function(a) {
  res <- a/43560
  return(res)
}
```

We define `sqft.to.acres` by assigning it to the output of `function`.
The list of argument names are contained within parentheses.
Next, the [body](reference.html#function-body) of the function--the statements that are executed when it runs--is contained within curly braces (`{}`).
The statements in the body are indented by two spaces.
This makes the code easier to read but does not affect how the code operates.

When we call the function, the values we pass to it are assigned to those variables so that we can use them inside the function.
Inside the function, we use a [return statement](reference.html#return-statement) to send a result back to whoever asked for it.

> ## Tip {.callout}
>
> One feature unique to R is that the return statement is not required.
> R automatically returns whichever variable is on the last line of the body
> of the function. Since we are just learning, we will explicitly define the
> return statement.

Let's try running our function.
Calling our own function is no different from calling any other function:

```{r}
sqft.to.acres(43560)
tmp <- 10000
sqft.to.acres(tmp)
```


## Combining functions

The real power of functions comes from mixing, matching and combining them
into ever large chunks to get the effect we want.

Here we convert sqft to acres and append it as an additional column of the dataset:

```{r}
add.acres <- function(dat) {
  myacres <- sqft.to.acres(dat$sqft)
  res <- cbind(dat, acres=myacres)
  return(res)
}
add.acres(head(lu))
```

Now we're going to define
a function that returns cities with sqft per job ratio above certain threshold:

```{r}
job.space.above <- function(dat, threshold=700) {
  idx <- which(dat$jobs > 0)
  ratio <- dat$non_res_sqft[idx]/dat$jobs[idx]
  return(dat[idx,]$city_name[ratio > threshold])
}
job.space.above(lu)
job.space.above(lu, 2000)
```


If you've been writing these functions down into a separate R script
(a good idea!), you can load in the functions into our R session by using the
`source` function:

```{r, eval=FALSE}
source("functions-lesson.R")
```

We've set a
*default argument* to `700` using the `=` operator
in the function definition. This means that this argument will
take on that value unless the user specifies otherwise.


> ## Tip: Pass by value {.callout}
>
> Functions in R almost always make copies of the data to operate on
> inside of a function body. When we modify `dat` inside the function
> we are modifying the copy of the gapminder dataset stored in `dat`,
> not the original variable we gave as the first argument.
>
> This is called "pass-by-value" and it makes writing code much safer:
> you can always be sure that whatever changes you make within the
> body of the function, stay inside the body of the function.
>

> ## Tip: Function scope {.callout}
>
> Another important concept is scoping: any variables (or functions!) you
> create or modify inside the body of a function only exist for the lifetime
> of the function's execution. When we call `cities.above`, the variables `dat`,
> `threshold`, `idx` and `ratio` only exist inside the body of the function. Even if we
> have variables of the same name in our interactive R session, they are
> not modified in any way when executing a function.
>


> ## Challenge 1 {.challenge}
>
> The `paste` function can be used to combine text together, e.g:
>
> ```{r}
> best_practice <- c("Write", "programs", "for", "people", "not", "computers")
> paste(best_practice, collapse=" ")
> ```
>
>  Write a function called `fence` that takes two vectors as arguments, called
> `text` and `wrapper`, and prints out the text wrapped with the `wrapper`:
>
> ```{r, eval=FALSE}
> fence(text=best_practice, wrapper="***")
> ```
>
> *Note:* the `paste` function has an argument called `sep`, which specifies the
> separator between text. The default is a space: " ". The default for `paste0`
> is no space "".
>

> ## Tip {.callout}
>
> R has some unique aspects that can be exploited when performing
> more complicated operations. We will not be writing anything that requires
> knowledge of these more advanced concepts. In the future when you are
> comfortable writing functions in R, you can learn more by reading the
> [R Language Manual][man] or this [chapter][] from
> [Advanced R Programming][adv-r] by Hadley Wickham. For context, R uses the
> terminology "environments" instead of frames.

[man]: http://cran.r-project.org/doc/manuals/r-release/R-lang.html#Environment-objects
[chapter]: http://adv-r.had.co.nz/Environments.html
[adv-r]: http://adv-r.had.co.nz/

## Control Flow

The meaning of the control flow structures `if else`, `for` and `while` is the same as in Python or other programming languages. What differs is the syntax. Let's learn these construct on examples.

First let's modify our `job.space.above` funtion to control for both ends of the distribution:
```{r}
job.space.outliers <- function(dat, threshold=700, above=TRUE) {
  idx <- which(dat$jobs > 0)
  ratio <- dat$non_res_sqft[idx]/dat$jobs[idx]
  if(above) {
  	res <- dat[idx,]$city_name[ratio > threshold]
  } else {
  	res <- dat[idx,]$city_name[ratio <= threshold]
  }
  return(res)
}
outl.above <- job.space.outliers(lu, 2000)
outl.above
outl.below <- job.space.outliers(lu, 50, above=FALSE)
outl.below 
```

We have our vectors of cities obtained `outl.above` and `outl.below` and we want to extract data corresponding to those cities from the households dataset:
```{r}
for(city in outl.above) {
	idx <- which(hh$city_name == city)
	print(hh[idx,])
}
```
This does not look very pretty. We can improve by binding the rows together:
```{r}
res <- NULL
for(city in outl.above) {
	idx <- which(hh$city_name == city)
	res <- rbind(res, hh[idx,])
}
print(res)
res <- NULL
for(city in outl.below) {
	idx <- which(hh$city_name == city)
	res <- rbind(res, hh[idx,])
}
print(res)
```

Sometimes you will find yourself needing to repeat an operation until a certain
condition is met. You can do this with a `while` loop.

Say we want the above data frame but with maximum of five rows:
```{r}
res <- NULL
i <- 1
while(i <= 5) {
	idx <- which(hh$city_name == outl.below[i])
	res <- rbind(res, hh[idx,])
	i = i + 1
}
print(res)
```

> ## Challenge 2 {.challenge}
>
> How would you vectorise the following loop
> 
> ```{r}
> res <- NULL
> for(city in outl.below) {
>   idx <- which(hh$city_name == city)
>   res <- rbind(res, hh[idx,])
> }
> ```

## Challenge solutions


> ## Solution to challenge 1 {.challenge}
>
>  Write a function called `fence` that takes two vectors as arguments, called
> `text` and `wrapper`, and prints out the text wrapped with the `wrapper`:
>
> ```{r}
> fence <- function(text, wrapper){
>   text <- c(wrapper, text, wrapper)
>   result <- paste(text, collapse = " ")
>   return(result)
> }
> best_practice <- c("Write", "programs", "for", "people", "not", "computers")
> fence(text=best_practice, wrapper="***")
> ```

> ## Solution to challenge 2 {.challenge}
>
> ```{r}
> hh[which(hh$city_name %in% outl.below),]
> ```
