---
layout: page
title: R for Data Analysis
subtitle: Homework
---

```{r, include=FALSE}
source("tools/chunk-options.R")
# Silently load in the data so the rest of the lesson works
lu <- read.table(file="data/city__dataset_table__lu_indicators__2010.tab", header=TRUE, sep="\t")
cities <- read.table(file="data/cities.csv", header=TRUE, sep=",", stringsAsFactors=FALSE)
lu <- merge(lu, cities, by="city_id", all=TRUE)
counties <- data.frame(county_id=1:4, county_name=c("Snohomish", "King", "Kitsap", "Pierce"))
lu <- merge(lu, counties, by="county_id", all=TRUE)
lu <- na.omit(lu)
```

This homework combines everything we learned: reading data from a file, transforming and adding columns, subsetting, plotting, reading help files.

Write a script that reads the land use dataset `city__dataset_table__lu_indicators__2010.tab`, joins it with cities and counties and for each county it creates a histogram of the non-residential sqft per job ratio. The histogram should have 20 bins. It should also include a thick red vertical line at the mean value of the ratio per county. Your result should look like this:

```{r, echo=FALSE, eval=TRUE}
lu <- cbind(lu, sqft_per_job=lu$non_res_sqft/lu$jobs)
par(mfrow=c(2,2))
for(county in 1:4){
	county.data <- subset(lu, county_id==county)
	hist(county.data$sqft_per_job, breaks=20, main=subset(counties, county_id==county)$county_name, xlab="sqft per job")
	abline(v=mean(county.data$sqft_per_job, na.rm=TRUE), col="red", lwd=2)
}
```

Hint: See `?hist` to find out how to control the number of bins in a histogram. Also for the title, you can extract the county name by subsetting the dataset `counties`.